---
title: "STA302 Final Project"
author: "Yixing"
date: "12/3/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("MASS")
library(MASS)
library(tidyverse)
library(dplyr)


```


## Cleaning and Processing Data

Isolating the chosen predictor variables "Adult Mortality", "Hepatitis B", "HIV Cases", "GDP per capita", "Life Expectancy", "Developed" 

```{r Project Part 1}
data <-read.csv("Life-Expectancy-Data-Updated.csv")
data <- subset(as.data.frame(data),
                   select =c(Region, Life_expectancy, Adult_mortality,Hepatitis_B,Incidents_HIV,GDP_per_capita,Economy_status_Developed))
write.csv(data, file = "Life-Expectancy-Data-Updated302.csv")
glimpse(data)
```
## Multiple Linear Model

The following variables are included in the Multiple Linear regression model: Life_expectancy as the response variable and Incidents_HIV, GDP_per_capita, Hepatitis_B, Adult-mortality and Economy_status_Developed as the predictor variables.

```{r}
model <- lm(Life_expectancy ~ GDP_per_capita + Adult_mortality + Incidents_HIV + Hepatitis_B + Economy_status_Developed, data = data)
summary(model)
```

### Extracting fitted values from the model

```{r}
predicted_life_expectancy <- fitted(model)
```

## Check conditions

### Check conditional mean response 

Check the mean responses are a single function of a linear combination involving beta by plotting a scatterplot of Life Expectancy versus fitted values

```{r}
# plot to check condition 1
plot(predicted_life_expectancy, data$Life_expectancy, main="Life Expectancy vs Fitted", xlab="Fitted", ylab="Life Expectancy")
abline(a = 0, b = 1, lty=2)
```
Since there is a general random diagonal scatter, we can conclude that Condition 1 holds. Now we will evaluate Condition 2:

### Check conditional mean predictors

```{r}
# plot to check condition 2
pairs(data[,3:7])
```
There are no clear non-linear patterns in these pairwise plots, so condition 2 holds.

## Residual Analysis

### Extracting residuals

```{r}
residuals <- resid(model)
```
### Residual Plots

To understand our model more and check for model assumptions the following plots:

```{r}
par(mfrow=c(2,3))
plot(residuals ~ predicted_life_expectancy, main = "Residuals vs Fitted", xlab = "Fitted", ylab = "Residual")
plot(residuals ~ data$Adult_mortality, main = "Residuals vs Adult mortatilty", xlab = "Adult mortality", ylab = "Residual")
plot(residuals ~ data$Hepatitis_B, main = "Residuals vs Hepatitis B", xlab = "Hepatitis B", ylab = "Residual")
plot(residuals ~ data$Incidents_HIV, main = "Residuals vs HIV prevalence", xlab = "HIV prevalence", ylab = "Residual")
plot(residuals ~ data$GDP_per_capita, main = "Residuals vs GDP per capita", xlab = "GDP per capita", ylab = "Residual")
boxplot(residuals ~ data$Economy_status_Developed, main = "Residuals vs Economy status", xlab = "Economy Status", ylab = "Residual",names =c("Not developed", "Developed"))
```
We can see in all the plots there are violations of linearity and constant variance, as there is a fanning pattern in nearly all of them and also a slight logarithmic curve in the Residuals vs GDP per capita.

```{r}
qqnorm(residuals)
qqline(residuals)
```
Very minimal deviation from the diagonal is not strong enough evidence of a violation of normality, and thus normality holds.

##Find an appropriate transformation

Evaluate the histogram of our response variable, Life Expectancy.

```{r}
hist(data$Life_expectancy, main = "Histogram of Life Expectancy", xlab = "Life Expectancy")
```
Since there is a left skew, and not a right skew, we have further reasont to use
a power transformation instead.

###Trying Box Cox Transformation on Y
```{r}
boxcox_val <- boxcox(model)
lambda <- boxcox_val$x[which.max(boxcox_val$y)]
lambda

model3 <- lm((data$Life_expectancy)^2 ~
               (data$Adult_mortality)+
               (data$Hepatitis_B) + 
               (data$Incidents_HIV) +
               (data$GDP_per_capita) +
               Economy_status_Developed, 
             data=data)

summary(model3)
```

Since Lambda is 1.8, it can be rounded to 2

###Rechecking assumptions for Box Cox Transformation on Y

```{r}
predicted_m3 <- fitted(model3)
residuals_m3 <- resid(model3)

par(mfrow=c(2,3))
plot(residuals_m3 ~ predicted_m3, main = "Residuals vs Box Cox Fitted", xlab = "Box Cox Fitted", ylab = "Residual")
plot(residuals_m3 ~ log(data$Adult_mortality), main = "Residuals vs Adult mortatilty", xlab = "Adult mortality", ylab = "Residual")
plot(residuals_m3 ~ log(data$Hepatitis_B), main = "Residuals vs Hepatitis B", xlab = "lnHepatitis B", ylab = "Residual")
plot(residuals_m3 ~ log(data$Incidents_HIV), main = "Residuals vs HIV prevalence", xlab = "HIV prevalence", ylab = "Residual")
plot(residuals_m3 ~ log(data$GDP_per_capita), main = "Residuals vs GDP per capita", xlab = "GDP per capita", ylab = "Residual")
boxplot(residuals_m3 ~ data$Economy_status_Developed, main = "Residuals vs Economy status", xlab = "Economy Status", ylab = "Residual",names =c("Not developed", "Developed"))

hist((data$Life_expectancy)^2, main = "Histogram of Box Cox Transformed Life Expectancy", xlab = "Box Cox Transformed Life Expectancy")
```

###Trying Box Cox Transformation on X
```{r}
p <- powerTransform(cbind(data[,3:7]))
summary(p)

#check the rounded powers in the summary and then plug in place of [lambda]
model2 <- lm(data$Life_expectancy ~
               (data$Adult_mortality)^lambda + #TODO: replace here
               (data$Hepatitis_B)^lambda + #TODO: replace here
               (data$Incidents_HIV)^lambda + #TODO:replace here
               (data$GDP_per_capita)^lambda + #TODO: replace here
               Economy_status_Developed, 
             data=data)

summary(model2)
```
###Rechecking assumptions for Box Cox Transformation on X

```{r}
predicted_m2 <- fitted(model2)
residuals_m2 <- resid(model2)

par(mfrow=c(2,3))
plot(residuals_m2 ~ predicted_m2, main = "Residuals vs Fitted", xlab = "Fitted", ylab = "Residual")
plot(residuals_m2 ~ log(data$Adult_mortality), main = "Residuals vs Adult mortatilty", xlab = "Box Cox Adult mortality", ylab = "Residual")
plot(residuals_m2 ~ log(data$Hepatitis_B), main = "Residuals vs Hepatitis B", xlab = "Box Cox Hepatitis B", ylab = "Residual")
plot(residuals_m2 ~ log(data$Incidents_HIV), main = "Residuals vs HIV prevalence", xlab = "Box Cox HIV prevalence", ylab = "Residual")
plot(residuals_m2 ~ log(data$GDP_per_capita), main = "Residuals vs GDP per capita", xlab = "Box Cox GDP per capita", ylab = "Residual")
boxplot(residuals_m2 ~ data$Economy_status_Developed, main = "Residuals vs Economy status", xlab = "Economy Status", ylab = "Residual",names =c("Not developed", "Developed"))

```

##Calculate Intervals using Inferential Tools

Our null hypothesis:
GDP_per_capita = 0
Adult_mortality = 0
Incidents_HIV = 0
Hepatitis_B = 0
Economy_status_Developed = 0

### Test Statistic
```{r}
summary(model)
```
Based on all our intercepts, our p-values are less than our significant value of 0.5 and therefore we 
can conclude there is a statistically significant linear relationship and we can reject our null hypothesis.

We can also see that our t values are greater than our critical value of 1.962.

###95% Prediction Interval
```{r}
# create new data to predict at
newdata <- data.frame(GDP_per_capita=5000, Adult_mortality=100, Hepatitis_B=85, Incidents_HIV=0.5, Economy_status_Developed=0)

predict(model, newdata = newdata, level = 0.95, interval = "prediction")
```


###95% Confidence Interval
```{r}
confint(model, level = 0.95)
```


##Test Dataset using Inferential Tools

###Calculate the decomposition of the sum of squares
